# Week 2 - Regression and model validation

*Describe the work you have done this week and summarize your learning.*

1. Read the data
```{r}
rm(list = ls()) # clear workspace first
learning2014 <- read.csv("~/Documents/GitHub/IODS-project/learning2014.csv") # my own data wrangling
# read.table("~/Documents/GitHub/IODS-project/data/learning2014.txt", sep = ",") # Kimmo's data
```

exploring the structure and dimensions of the dataset
```{r}
str(learning2014)
dim(learning2014)
```

The data includes 7 variables and 166 observations. 
The variables are: 
1. Gender (Female = 1  Male = 2)
2. Age
3. Attitude
4-6. Means of the deep, strategic and surface learning scores
7. Exam points


```{r}
library(ggplot2) # Access the gglot2 library
```

2. Show a graphical overview of the data. 
```{r}
# show gender distributions as bar graph
p1 <- ggplot(learning2014, aes(gender))
p1 + geom_bar()

# display variable distributions as histogram
p2 <- ggplot(learning2014, aes(Age))
p2 + geom_histogram(binwidth = 5)

p3 <- ggplot(learning2014, aes(Attitude))
p3 + geom_histogram(binwidth = 2)

p4 <- ggplot(learning2014, aes(deep))
p4 + geom_histogram(binwidth = 0.5)

p5 <- ggplot(learning2014, aes(stra))
p5 + geom_histogram(binwidth = 0.5)

p6 <- ggplot(learning2014, aes(surf))
p6 + geom_histogram(binwidth = 0.5)

p7 <- ggplot(learning2014, aes(Points))
p7 + geom_histogram(binwidth = 2)



# show relationships between variables 
p8 <- ggplot(learning2014, aes(x = Attitude, y = Points, col=gender))
p8 + geom_point() + ggtitle("Relationship between exam points and deep learning") + geom_smooth(method = "lm")

p9 <- ggplot(learning2014, aes(x = deep, y = Points, col=gender))
p9 + geom_point() + ggtitle("Relationship between exam points and deep learning") + geom_smooth(method = "lm")

p10 <- ggplot(learning2014, aes(x = stra, y = Points, col=gender))
p10 + geom_point() + ggtitle("Relationship between exam points and strategic learning") + geom_smooth(method = "lm")

p11 <- ggplot(learning2014, aes(x = surf, y = Points, col=gender))
p11 + geom_point() + ggtitle("Relationship between exam points and surface learning") + geom_smooth(method = "lm")

p12 <- ggplot(learning2014, aes(x = Age, y = Points, col=gender))
p12 + geom_point() + ggtitle("Relationship between age and exam points") + geom_smooth(method = "lm")

p13 <- ggplot(learning2014, aes(x = Age, y = Attitude, col=gender))
p13 + geom_point() + ggtitle("Relationship between age and attitudes") + geom_smooth(method = "lm")


```
```{r}
library(GGally)
pairs(learning2014[-1], col = learning2014$gender)

p <- ggpairs(learning2014, mapping = aes(col=gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))

# draw the plot
p
```

Summary of the variables
```{r}
summary(learning2014)

# another way of summarising the variables
library(dplyr)
learning2014 %>%
  group_by(gender) %>%
  summarise(mean = mean(Attitude), n = n())
```


Describe and interpret the outputs, commenting on the distributions of the variables and the relationships between them. 

3. Fitting of a regression model to study whether the three explanatory variables (the learning strategies) explain the exam points. The learning strategies (deep, strategic or surface) did not explain the exam points significantly.
```{r}
# fit a linear model
my_model1 <- lm(Points ~ deep + stra + surf, data = learning2014) # how to use three explanatory vars
summary(my_model1)

my_model2 <- lm(Points ~ stra + surf, data = learning2014) 
summary(my_model2)

my_model3 <- lm(Points ~ surf, data = learning2014) 
summary(my_model3)


```
Produce the following diagnostic plots: Residuals vs Fitted values, Normal QQ-plot and Residuals vs Leverage. 
```{r}
plot(my_model1)
```

