# Week 2 - Regression and model validation

*Read the data*
```{r}
rm(list = ls()) # clear workspace first
learning2014 <- read.csv("~/Documents/GitHub/IODS-project/learning2014.csv") # my own data wrangling
# read.table("~/Documents/GitHub/IODS-project/data/learning2014.txt", sep = ",") # Kimmo's data
```

*Explore the structure and dimensions of the dataset*
```{r}
str(learning2014)
dim(learning2014)
```

The data includes 7 variables and 166 observations. 
The variables are: 
1. Gender (Female = 1  Male = 2)
2. Age
3. Attitude
4-6. Mean scores of the deep, strategic and surface learning 
7. Exam points


```{r}
library(ggplot2) # Access the gglot2 library
```

*Show a graphical overview of the data*
```{r}
# show gender distributions as bar graph
p1 <- ggplot(learning2014, aes(gender))
p1 + geom_bar()

# display variable distributions as histogram
p2 <- ggplot(learning2014, aes(Age))
p2 + geom_histogram(binwidth = 5)

p3 <- ggplot(learning2014, aes(Attitude))
p3 + geom_histogram(binwidth = 2)

p4 <- ggplot(learning2014, aes(deep))
p4 + geom_histogram(binwidth = 0.5)

p5 <- ggplot(learning2014, aes(stra))
p5 + geom_histogram(binwidth = 0.5)

p6 <- ggplot(learning2014, aes(surf))
p6 + geom_histogram(binwidth = 0.5)

p7 <- ggplot(learning2014, aes(Points))
p7 + geom_histogram(binwidth = 2)



# show relationships between variables 
p8 <- ggplot(learning2014, aes(x = Attitude, y = Points, col=gender))
p8 + geom_point() + ggtitle("Relationship between exam points and deep learning") + geom_smooth(method = "lm")

p9 <- ggplot(learning2014, aes(x = deep, y = Points, col=gender))
p9 + geom_point() + ggtitle("Relationship between exam points and deep learning") + geom_smooth(method = "lm")

p10 <- ggplot(learning2014, aes(x = stra, y = Points, col=gender))
p10 + geom_point() + ggtitle("Relationship between exam points and strategic learning") + geom_smooth(method = "lm")

p11 <- ggplot(learning2014, aes(x = surf, y = Points, col=gender))
p11 + geom_point() + ggtitle("Relationship between exam points and surface learning") + geom_smooth(method = "lm")

p12 <- ggplot(learning2014, aes(x = Age, y = Points, col=gender))
p12 + geom_point() + ggtitle("Relationship between age and exam points") + geom_smooth(method = "lm")

p13 <- ggplot(learning2014, aes(x = Age, y = Attitude, col=gender))
p13 + geom_point() + ggtitle("Relationship between age and attitudes") + geom_smooth(method = "lm")


```
```{r}
library(GGally)
pairs(learning2014[-1], col = learning2014$gender)

p <- ggpairs(learning2014, mapping = aes(col=gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))

# draw the plot
p
```

*Summary of the variables*
```{r}
summary(learning2014)

# another way of summarising the variables
#library(dplyr)
#learning2014 %>%
#  group_by(gender) %>%
#  summarise(mean = mean(Attitude), n = n())
```

The sample consist of mainly female participants.
The majority of participants are between 20 to 30 years old.
The variables are mostly normally distributed.
The attitudes predict the exam points, but the learning strategies do not explain the exam points.
Age does not explain the exam points or attitudes.


*Fitting of a regression model to study whether attitudes, strategic and surface learning strategies explain the exam points. The learning strategies (strategic and surface) did not explain the exam points significantly. The exam points were significantly (p = 4.12e-09) explained by attitudes*
```{r}
# fit a linear model
my_model1 <- lm(Points ~ Attitude + stra + surf, data = learning2014) # how to use three explanatory vars
summary(my_model1)

my_model2 <- lm(Points ~ Attitude, data = learning2014) 
summary(my_model2)

```
*Produce the diagnostic plots: Residuals vs Fitted values, Normal QQ-plot and Residuals vs Leverage*
```{r}
plot(my_model2)
```

